### 2017-02-07 MSR-Scratch.tex
Paper structure

\documentclass[10pt, conference]{IEEEtran}

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}

\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{color}

\newcommand{\fenia}[1]{\emph{\color{blue}Fenia says: #1}}
\newcommand{\felienne}[1]{\emph{\color{green}Felienne says: #1}}
\newcommand{\grex}[1]{\emph{\color{yellow}Gregorio says: #1}}
\newcommand{\jemole}[1]{\emph{\color{red}Jesús says: #1}}

\begin{document}

\title{A dataset of Scratch programs,\\Scraped, Shaped and Graded}

\author{
	\IEEEauthorblockN{Efthimia Aivaloglou\IEEEauthorrefmark{1}, Felienne Hermans\IEEEauthorrefmark{1}, Gregorio Robles\IEEEauthorrefmark{2}, and Jes{\'u}s Moreno-Le{\'o}n\IEEEauthorrefmark{3}}

	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
			Email: \{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
	\IEEEauthorblockA{\IEEEauthorrefmark{2}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
		Email: grex@gsyc.urjc.es}
	\IEEEauthorblockA{\IEEEauthorrefmark{3}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
		Email: jesus.moreno@programamos.es}
}

\maketitle

\begin{abstract}
In this paper, we present a collection of 250K Scratch programs scraped from the Scratch project repository. We processed the program data to encode it into a database that facilitates querying and further analysis. The dataset is intended as an input for research on computing education and source code analysis.
\end{abstract}

\begin{IEEEkeywords}
Scratch; dataset;
\end{IEEEkeywords}

\section{Introduction}
\cite{Aivaloglou_2016}
\cite{Robles_2017}

\section{Dataset Construction}
a description of the methodology used to gather it (preferably with the tool used to create/generate the data)

\section{Data Representation}
a description of the data source
a description of the storage mechanism, including a schema if applicable,

\section{Using the Dataset / Enabled Research / Research Opportunities}
a description of how the data has been used by others,

\section{Extending the Dataset}
ideas for what future research questions could be answered or what further improvements could be made to the data set

\section{Limitations}
any limitations and/or challenges in creating or using this data set.

\section{Conclusion}

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,ScratchDataset}

\end{document}

### 2017-02-07 MSR-Scratch.tex
Introduction

@@
 
+\usepackage{url}
 \usepackage{graphicx}
@@
 \usepackage{color}
+
+\newcommand{\nPrograms}{250,163}
+\newcommand{\nAnalyzedPrograms}{247,798}
+\newcommand{\nemptyPrograms}{14,307}
+\newcommand{\nScriptPrograms}{233,491}
+\newcommand{\nLOC}{36,085,654}
+\newcommand{\nscripts}{4,049,356}
 
@@
 \section{Introduction}
-\cite{Aivaloglou_2016}
-\cite{Robles_2017}
+Scratch \cite{resnick_scratch:_2009} is a block-based programming language developed to serve as a stepping stone for children from 8 to 16 years old to the more advanced world of computer programming.
+It offers a web-based programing environment that enables creating games and interactive animations. The public repository of Scratch programs contains over 19 million projects and 16 million users.
+
+A number of works in the computing education research field attempt to assess the programming skills that novice programmers develop in the Scratch environment.
+Some utilize program data collected during specific programming courses (e.g., \cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab, which contains data for Scratch projects created until 2012 (e.g., \cite{fields_2014, yang_2015, Dasgupta_2016}).
+In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices within Scratch programs \cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g.
+Hairball \cite{boe_hairball:_2013} and Dr.
+Scratch \cite{moreno_automatic_2014}).
+
+While Scratch is receiving increasing interest as an introductory programming language, there is no recent dataset of Scratch programs available to the research community.
+The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the Scratch 2 web programming interface in 2013.
+It is since then that the popularity of Scratch started increasing\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
+
+The goal of this paper is to present an open and timely dataset of recent Scratch programs, along with their metadata, that can facilitate quantitative research in the fields of source code analysis and computing education.
+The dataset contains 250 thousand Scratch projects, from 100 thousand different users, that were scraped from the Scratch project repository.
+It is made available as a database\footnote{{add anonymized address}} which includes, for each Scratch project, its metadata and the program data, along with project grading results from the Dr.
+Scratch quality assessment tool \cite{moreno_automatic_2014}.
 

### 2017-02-07 MSR-Scratch.tex
Initial input on dataset construction

@@
 \section{Dataset Construction}
+\label{dataset}
 a description of the methodology used to gather it (preferably with the tool used to create/generate the data)
+
+The dataset that we used for this study is the one made available from~\cite{Aivaloglou_2016}.
+The dataset consists of 250,166 projects scraped from the Scratch project repository.
+Those are the projects most recently shared in the Scratch projects page\footnote{\url{https://scratch.mit.edu/explore/projects/all/}} when the web scraping program run on March 2nd 2016.
+The web scraping program, as well as all scraped projects in JSON format are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}
+
+In order to collect the set of Scratch projects in the dataset we built a web scraping program.
+The scraping program, called Kragle, starts by reading the Scratch projects page\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} and thus obtains the project identifiers of projects that were most recently shared.
+Subsequently, Kragle retrieves the JSON code for each of the listed projects.
+
+% For a given id $x$ a Scratch programs JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}.
+Does not fit on the page nicely and might also not be needed?
+
+We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained a little over 250,000 projects.
+Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
+Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}
+
+Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
+This resulted in a list of used blocks per project, with the sprites and the stage of the project.
+We also cross referenced all blocks with the Scratch wiki to determine the shapes and the category of all blocks.
+For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
+
+\subsection{Data analysis}
+\label{dataAnalysis}
+All scraped project data, including the list of used blocks and parameters, were imported in a relational database.
+We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions.
+We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects.
+Data for these cases are provided as part of the dataset.\footnotemark[\ref{repo}]
 

### 2017-02-07 MSR-Scratch.tex
Dataset Construction

--- 2017-02-07 MSR-Scratch.tex
@@
 
-\title{A dataset of Scratch programs,\\Scraped, Shaped and Graded}
+\title{A dataset of Scratch programs,\\Scraped, Shaped and Scored}
 
@@
 The dataset contains 250 thousand Scratch projects, from 100 thousand different users, that were scraped from the Scratch project repository.
-It is made available as a database\footnote{{add anonymized address}} which includes, for each Scratch project, its metadata and the program data, along with project grading results from the Dr.
+It is made available as a database\footnote{{add anonymized address}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
 Scratch quality assessment tool \cite{moreno_automatic_2014}.
@@
 \section{Dataset Construction}
-a description of the methodology used to gather it (preferably with the tool used to create/generate the data)
+\label{dataset}
+
+\subsection{Data collection}
+To collect the data from the web interface of the Scratch project repository we built a scraping program.
+The web scraping program, called Kragle, starts by reading the Scratch projects page\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} and thus obtains the project identifiers of projects that were most recently shared.
+Subsequently, Kragle retrieves the JSON code for each of the listed projects\footnote{For a given project id $x$, the program's JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}}.
+
+We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON code for \nPrograms projects.
+Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
+Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}
+
+Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
+This resulted in a list of used blocks per project, within the sprites and the stage of the project.
+We also cross referenced all blocks with the Scratch wiki to determine the shapes and the category of all blocks.
+For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
+
+\subsection{Calculation of programming mastery scores}
+\fenia{Description of how the mastery metrics for       
+	[Abstraction]
+	,[Parallelism]
+	,[Logic]
+	,[Synchronization]
+	,[FlowControl]
+	,[UserInteractivity]
+	,[DataRepresentation]
+	,[Mastery]
+	,[Clones]
+	,[CustomBlocks]
+	,[InstancesSprites]
+	were calculated}
+
+\subsection{Importing the data}
+\label{dataAnalysis}
+All scraped project data and metadata, including the list of used blocks and parameters, were imported in a relational database.
+We also imported the data on the shapes and the categories of the Scratch blocks.
+We then used SQL queries for normalizing the data and bringing it in its final schema, which is drawn in Figure???
 

### 2017-02-09 MSR-Scratch.tex
DB Schema table

@@
 
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{llp{5.4cm}}
+		\textbf{Table}& \textbf{Key} & \textbf{Attribute(Description)}\\
+		\hline
+		\textbf{\texttt{project}} & PK & \texttt{p\_id} (Scratch project ID)\\
+		&  & 							\texttt{project-name} (name given to project)\\
+		&  & 							\texttt{username} (author's Scratch 
+		username)\\
+		&  & 							\texttt{total-views} (project views number)\\
+		&  & 							\texttt{total-remixes} (project remixes number)\\
+		&  & 							\texttt{total-favorites} (total users favoriting)\\
+		&  & 							\texttt{total-loves} (total users `loving' the project)\\
+		& & 							\texttt{is-remix} (calculated column, true if project is a remix of another one)\\
+		\hline
+		\textbf{\texttt{script}} & PK & \texttt{script\_id} (auto increment)\\
+		& FK & 							\texttt{project\_id} (\scriptsize{\texttt{*\ldots1 project:p\_id}})\\
+		& & 							\texttt{sprite-type} (one of [sprite, stage, procDef])\\
+		& & 							\texttt{sprite-name} (name sprite the script is in)\\
+		& & 							\texttt{script-rank} (project-level ranking of script)\\
+		& & 							\texttt{coordinates} (x-y location of script in Scratch editor)\\
+		& & 							\texttt{total-blocks} (calculated column, number of blocks comprising the script)\\
+		\hline
+		\textbf{\texttt{procedure}} & PK,FK & \texttt{script\_id} (\scriptsize{\texttt{1\ldots1 script:script\_id}})\\
+		& & 							\texttt{proc-name} (name given to custom block)\\
+		& & 							\texttt{total-args} (number of input arguments)\\
+		\hline
+		\textbf{\texttt{block}} & PK,FK  & \texttt{script\_id} (\scriptsize{\texttt{*\ldots1 script:script\_id}})\\
+		& PK & \texttt{block-rank} (script-level ranking of block)\\
+		& FK & \texttt{block-type} (\scriptsize{\texttt{*\ldots1 blockType:b-type}})\\
+		&  & \texttt{parameter-1} (value of 1st input parameter)\\
+		&  &\vdots\\
+		&  & \texttt{parameter-24} (value of 24th input parameter)\\
+		\hline
+		\textbf{\texttt{blockType}} & PK & \texttt{b-type} (Scratch name of predefined block)\\
+		&  & \texttt{category} (Scratch block category)\\
+		&  & \texttt{shape} (One of the 8 block shapes in the Scratch interface)\\
+		&  & \texttt{is-input} (true if the block receives user input)\\
+		\hline
+		\textbf{\texttt{remix}} & PK  & \texttt{remix\_id} (\texttt{\scriptsize{project\_id}} of created project)\\
+		&  & \texttt{from-project\_id} (\texttt{\scriptsize{project\_id}} of original project)\\
+		\hline
+		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 project:p\_id}})\\
+		& & \texttt{Abstraction} ()\\
+		& & \texttt{Parallelism} ()\\
+		& & \texttt{Logic} ()\\
+		& & \texttt{Synchronization} ()\\
+		& & \texttt{FlowControl} ()\\
+		& & \texttt{UserInteractivity} ()\\
+		& & \texttt{DataRepresentation} ()\\
+		& & \texttt{Mastery} ()\\
+		& & \texttt{Clones} ()\\
+		& & \texttt{CustomBlocks} ()\\
+		& & \texttt{InstancesSprites} ()\\
+		\hline
+	\end{tabular}
+	\caption{Database schema: Tables and Attributes}
+	\label{tbl-dbschema}
+\end{table}
+
 \section{Using the Dataset / Enabled Research / Research Opportunities}

### 2017-02-09 MSR-Scratch.tex
Added Scratch example

@@
 The dataset contains 250 thousand Scratch projects, from 100 thousand different users, that were scraped from the Scratch project repository.
-It is made available as a database\footnote{{add anonymized address}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
+It is made available as a database\footnote{{\fenia{add anonymized address, update it in repo footnote}}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
 Scratch quality assessment tool \cite{moreno_automatic_2014}.
@@
 We also imported the data on the shapes and the categories of the Scratch blocks.
-We then used SQL queries for normalizing the data and bringing it in its final schema, which is drawn in Figure???
-
-\section{Data Representation}
-a description of the data source
-a description of the storage mechanism, including a schema if applicable,
+We then used SQL queries for normalizing the data and bringing it in its final schema, described in Table \ref{tbl-dbschema}.
 
@@
 
+\section{Data Representation}
+
+add example scratch project
+add some statistics
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{scratchExample}
+	\caption{Example of a Scratch program with one sprite and three scripts}
+	\label{fig-scratchExample}
+\end{figure}
+
+\fenia{Description of Dr Scratch outputs in Table: grades}
+
 \section{Using the Dataset / Enabled Research / Research Opportunities}

### 2017-02-09 MSR-Scratch.tex
Size and summary statistics tables

@@
 
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{lr}
+		\hline
+		Projects & 250,163 \\
+		Non-empty projects & 233,491 \\
+		Remixes of other projects & 12,167 \\
+		Usernames & 109,960 \\
+		Scripts & 4,049,356 \\
+		Blocks & 36,085,654 \\
+		Custom block definitions & 209,769 \\
+		Projects analyzed by Dr. Scratch & 231,050 \\
+		\hline
+	\end{tabular}
+	\caption{Dataset contents}
+	\label{tbl-size}
+\end{table}
+
+\begin{table*}[ht]
+	\centering
+	\begin{tabular}{lrrrrrr}
+		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
+		\hline
+			Projects per username&2.27&1&1&1&2&868\\
+			Sprites per project&5.30&0&1&2&5&525\\
+			Scripts per project&16.19&0&2&4&11&3,038\\
+			Blocks per project&144.25&0&9&26&69&34,622\\
+			Custom block definitions per project&0.64&0&0&0&0&372\\
+			Total views per project&5.76&0&1&1&4&27,993\\
+			Total remixes per project&197.12&0&0&0&0&14,951,164\\
+			Total favorites per project&0.54&0&0&0&0&2,582\\
+			total-loves&0.41&0&0&0&0&1940\\
+			Dr. Scratch mastery score per project&8.92&0&6&8&11&21\\
+		\hline
+	\end{tabular}
+	\caption{Summary statistics of the dataset}
+	\label{tbl-stats}
+\end{table*}
+
 \fenia{Description of Dr Scratch outputs in Table: grades}

### 2017-02-10 MSR-Scratch.tex
Dataset description

@@
 The goal of this paper is to present an open and timely dataset of recent Scratch programs, along with their metadata, that can facilitate quantitative research in the fields of source code analysis and computing education.
-The dataset contains 250 thousand Scratch projects, from 100 thousand different users, that were scraped from the Scratch project repository.
+The dataset contains 250 thousand Scratch projects, from more than 100 thousand different users, that were scraped from the Scratch project repository.
 It is made available as a database\footnote{{\fenia{add anonymized address, update it in repo footnote}}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
@@
 For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
+We included the blocks from Scratch extensions, such as the LEGO WeDo extensions, that we found in the dataset.
 
@@
 We also imported the data on the shapes and the categories of the Scratch blocks.
-We then used SQL queries for normalizing the data and bringing it in its final schema, described in Table \ref{tbl-dbschema}.
+We then used SQL queries for normalizing the data and bringing it in its final schema, outlined in Table \ref{tbl-dbschema}.
 
-\begin{table}[ht]
+\begin{table}[]
 	\centering
@@
 		\textbf{\texttt{script}} & PK & \texttt{script\_id} (auto increment)\\
-		& FK & 							\texttt{project\_id} (\scriptsize{\texttt{*\ldots1 project:p\_id}})\\
+		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 project:p\_id}})\\
 		& & 							\texttt{sprite-type} (one of [sprite, stage, procDef])\\
@@
 		\hline
-		\textbf{\texttt{procedure}} & PK,FK & \texttt{script\_id} (\scriptsize{\texttt{1\ldots1 script:script\_id}})\\
+		\textbf{\texttt{procedure}} & PK,FK & \texttt{script\_id} (\texttt{\scriptsize{1\ldots1 script:script\_id}})\\
 		& & 							\texttt{proc-name} (name given to custom block)\\
@@
 		\hline
-		\textbf{\texttt{block}} & PK,FK  & \texttt{script\_id} (\scriptsize{\texttt{*\ldots1 script:script\_id}})\\
+		\textbf{\texttt{block}} & PK,FK  & \texttt{script\_id} (\texttt{\scriptsize{*\ldots1 script:script\_id}})\\
 		& PK & \texttt{block-rank} (script-level ranking of block)\\
-		& FK & \texttt{block-type} (\scriptsize{\texttt{*\ldots1 blockType:b-type}})\\
+		& FK & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockType:b-type}})\\
 		&  & \texttt{parameter-1} (value of 1st input parameter)\\
@@
 		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 project:p\_id}})\\
-		& & \texttt{Abstraction} ()\\
+		& & \texttt{Abstraction} (\fenia{short descriptions})\\
 		& & \texttt{Parallelism} ()\\
@@
 
-\section{Data Representation}
+\section{Dataset Description}
 
-add example scratch project
-add some statistics
+\subsection{Data Representation}
+
+The projects' data are stored in a relational database.
+Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
+If a project is a remix of another one, the original project can be found in the \textbf{\texttt{remix}} table.
+Table \textbf{\texttt{grades}} stores the Dr. Scratch results for the programming mastery metrics per project.
 
@@
  	\includegraphics[width=0.45\textwidth]{scratchExample}
-	\caption{Example of a Scratch program with one sprite and three scripts}
+ 	\caption{Example of a Scratch program with three scripts in the same spite}
  	\label{fig-scratchExample}
@@
 
-\begin{table}[ht]
+The schema of the database, outlined in Table \ref{tbl-dbschema}, reflects the structure of the Scratch programs.
+Projects contain sprites, which are entities with their own associated code.
+The code is organized into \textbf{\texttt{script}}s, i.e., groups of Scratch code blocks, each script belonging a sprite named \texttt{sprite-name}.
+In the example in Figure \ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
+Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedure}} table.
+
+The coding elements in the Scratch environment are the \textbf{\texttt{block}}s, which can receive input parameters.
+In the example in the figure, block `go to x:0 y:0' receives as input two numerical parameters.
+Custom clock invocations are stored in this table just like regular blocks.
+To cater for the varying number of input parameters that those blocks can have, this table can store up to the maximum number of input parameters found in the dataset, which is 24.
+The ordering of the blocks in the scripts can be retrieved using the \texttt{block-rank} field, which represents a depth-first ranking of the blocks.
+In the script of the `backflip' custom block definition in the example, the ranking of the blocks would be `define backflip', `set counter to', `+', `counter', followed by the rest of the blocks.
+The clocks can be of different shapes and categories.
+In table \textbf{\texttt{blockType}} we have stored the encoding of all types of Scratch blocks and the blocks from Scratch extensions.
+
+\subsection{Dataset Contents}
+
+\begin{table}[]
 	\centering
@@
 
-\begin{table*}[ht]
+\begin{table*}[]
 	\centering
@@
 		Total favorites per project&0.54&0&0&0&0&2,582\\
-			total-loves&0.41&0&0&0&0&1940\\
 		Dr. Scratch mastery score per project&8.92&0&6&8&11&21\\
@@
 
-\fenia{Description of Dr Scratch outputs in Table: grades}
+As shown in Table \ref{tbl-size}, the dataset contains the metadata of 250,163 Scratch projects and the code of 233,491 of those that are non-empty.
+The projects were created from 109,960 different users.
+Most of those, as shown in Table \ref{tbl-stats}, have created a single project.
+However, 40 thousand users have created more than one and up to 868 projects in the dataset.
+
+The majority of the projects have been viewed only once, and they have not been favorited or remixed.
+Around 10 thousand of the total projects have been remixed at least once, while the dataset includes cases of very popular projects with more than 100 remixes and even one thousand cases with more than 100 views.
+
+Table \ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21.
+The table also highlights the existence of surprisingly complex projects, with 525 sprites, 3 thousand scripts as 34 thousand blocks.
+
+\fenia{Should we add some stats about the Dr Scratch master scores?}
 

### 2017-02-10 MSR-Scratch.tex
Enabled research

@@
 
-\begin{table}[]
-	\centering
-	\begin{tabular}{lr}
-		\hline
-		Projects & 250,163 \\
-		Non-empty projects & 233,491 \\
-		Remixes of other projects & 12,167 \\
-		Usernames & 109,960 \\
-		Scripts & 4,049,356 \\
-		Blocks & 36,085,654 \\
-		Custom block definitions & 209,769 \\
-		Projects analyzed by Dr. Scratch & 231,050 \\
-		\hline
-	\end{tabular}
-	\caption{Dataset contents}
-	\label{tbl-size}
-\end{table}
-
 \begin{table*}[]
@@
 
+\begin{table}[]
+	\centering
+	\begin{tabular}{lr}
+		\hline
+		Projects & 250,163 \\
+		Non-empty projects & 233,491 \\
+		Remixes of other projects & 12,167 \\
+		Usernames & 109,960 \\
+		Scripts & 4,049,356 \\
+		Blocks & 36,085,654 \\
+		Custom block definitions & 209,769 \\
+		Projects analyzed by Dr. Scratch & 231,050 \\
+		\hline
+	\end{tabular}
+	\caption{Dataset contents}
+	\label{tbl-size}
+\end{table}
+
 As shown in Table \ref{tbl-size}, the dataset contains the metadata of 250,163 Scratch projects and the code of 233,491 of those that are non-empty.
@@
 
-\section{Using the Dataset / Enabled Research / Research Opportunities}
-a description of how the data has been used by others,
+\section{Enabled Research and Dataset Extension}
+\fenia{What about our own papers utilizing the dataset?
+(\cite{Aivaloglou_2016} and \cite{Robles_2017}).
+We have to anonymize the submission, so do we add them now or add a note that we omitted them and add them to the camera ready?}
 
-\section{Extending the Dataset}
-ideas for what future research questions could be answered or what further improvements could be made to the data set
+The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education is receiving significant research attention.
+A number of studies have been carried out during the last years on understanding of programming concepts and the programming practices of novice programmers in block-based environments, on the programming skills they develop, and on the quality of Scratch programs.
+
+The first research direction that this dataset can be utilized for concerns the \textit{assessment of the programming skills that novice programmers develop in the Scratch environment}.
+The dataset can be used for examining indications of learning of programming concepts and abstractions.
+Existing works in this research direction include the work by Maloney \emph{et al.} \cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers, and found that within the least utilized ones are boolean operators and variables.
+Another study on the internalization of programming concepts with Scratch with 46 students was presented in \cite{meerbaum-salant_learning_2010}.
+Concepts like loops, conditional loops, message passing, initialization, variables and concurency were examined, and it was found that students had problems with the last three.
+In \cite{wilson_evaluation_2012}, 29 projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
+
+Towards this research direction, the dataset makes the source code of the Scratch programs available for static analysis and can be used for quantitatively evaluating the application of programming concepts and abstractions through the presence of Scratch blocks of the corresponding types.
+
+Another promising research direction relates to the \textit{quality of the programming artifacts developed in the Scratch environment}.
+The dataset can be used for quality assessment, for the identification of programming smells, and for exploring indications of harmful programming habits.
+Existing works in this direction include a study in a classroom setting \cite{Meerbaum_habits_2011}, which highlighted two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming.
+The authors connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops.
+Related to smell detection are the Scratch automated quality analysis tools Hairball \cite{boe_hairball:_2013}, a lint-like static analysis tool for Scratch that can detect initialization problems and unmatched broadcast and receive blocks, and Dr.
+Scratch \cite{moreno-leon_dr._2015}, which extends Hairball to detect two bad programming habits in Scratch: not changing the default object names and duplicating scripts.
+
+This dataset will enable examining code quality and code smells on a large set of Scratch projects.
+Static analysis of the source code can also be performed for the identification of other types of smells \cite{fowler_refactoring:_1999} that might be common in the artifacts of novice programmers.
+
+Moving from the software engineering to the computing education research field, a research direction with increased data requirements concerns the \textit{learning progressions of novice programmers}.
+This includes examining the factors that support learning and improving their programming skills.
+Existing works in this area have used the dataset of Scratch programs created until 2012, before the current web-based programming environment was introduced, and described in the Introduction.
+They include the works of Dasgupta et al., who investigated how project remixing relates to the adoption of new computational thinking concepts \cite{Dasgupta_2016}, and Yang et al., who examined the learning patterns of programmers in terms of block use over their first 50 projects \cite{yang_2015}.
+
+Enabling research towards this direction requires extending the dataset with the complete set of projects for the included users, along with their creation dates.
+This would enable examining how the programming skills and mastery of the users evolved through time, possibly through remixing the projects of other users.
+An even richer extension of the dataset would be with, possibly daily, snapshots of the included projects that are still in development.
+This would enable reconstructing some project versioning information, not available from the Scratch web interface, and thus examining how the projects are developed and evolve over time.
 

### 2017-02-16 MSR-Scratch.tex
Limitations

@@
 The projects' data are stored in a relational database.
-Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
+Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{\label{fn-authorpage}Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
 If a project is a remix of another one, the original project can be found in the \textbf{\texttt{remix}} table.
@@
 \section{Limitations}
-any limitations and/or challenges in creating or using this data set.
+The only information about the authors of the Scratch projects contained in the dataset is their username.
+This suffices for creating project portfolios of the authors and facilitating research in the directions described in the previous section.
+However, other directions in the computing education research field require richer author data, and especially data on their gender and their age.
+There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman \cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelization, conditionals and, especially, variables were under-represented by all grades apart from 5 an 6.
+The effect of gender and account age were also examined in \cite{fields_2014} in relation to the use of programming concepts.
+Studies on the effect of parameters like the age are not only interesting but also of direct value to educators, since they give insights on when it is better to introduce computational thinking concepts in the curriculum of schools.
+However, this dataset does not include information on the gender and the age of authors and cannot be extended to do so using the current Scratch web interface because this information is not available in the user profile page\footnotemark[\ref{fn-authorpage}], even though it is provided by the users upon registration.
+
+Another limitation of this dataset is that we did not scrape a random sample of Scratch projects, but the 250,000 most recent ones during the time that we run the scraper.
+It could be the case that the programming habits of Scratch users are changing over time.
+However, we counterbalanced that by collecting a large dataset which comprises around 1.3\% of all 19 million shared Scratch projects, and is still the most recent one to be made available to the research community.
 

### 2017-02-16 MSR-Scratch.tex
Addition to enabled research

@@
 
+Apart from quantitative studies, the dataset can support the design of experiments and field studies, whose material often includes Scratch programs with specific characteristics.
+For example, in \cite{hermans_2016} examples of Scratch programs exhibiting specific smells were used in a controlled experiment to determine how the smells affected the understanding and the modification of the programs.
+This queryable dataset can support finding example programs with characteristics according to the experiment design and the field of the study.
+
 \section{Limitations}

### 2017-02-16 MSR-Scratch.tex
Conclusion

@@
 \section{Conclusion}
+The increasing popularity of Scratch as an introductory programming language makes it an increasingly popular research target in the computing education research field.
+This paper presented an open and timely dataset of recent Scratch programs, that can facilitate research in the fields of source code analysis and computing education.
+The dataset can support various research topics, including the ones related to the assessment of the programming skills that novice programmers develop in the Scratch environment and the exploration of their learning progressions, and issues related to the quality of the programs and the identification of smells and bad programming habits.
+
+The dataset contains 250 thousand Scratch projects, from more than 100 thousand different users, that were scraped from the Scratch project repository.
+It is made available as a database\footnote{{\fenia{add anonymized address, update it in repo footnote}}} which includes the source code of the Scratch projects, their metadata, and their programming mastery scoring results.
 

### 2017-02-16 MSR-Scratch.tex
abstract

@@
 \begin{abstract}
-In this paper, we present a collection of 250K Scratch programs scraped from the Scratch project repository.
+Scratch is increasingly popular, both as an introductory programming language and as a research target in the computing education research field.	
+In this paper, we present an open dataset of 250K recent Scratch programs from 100K authors scraped from the Scratch project repository.
 We processed the program data to encode it into a database that facilitates querying and further analysis.
+We further enriched the dataset with project metadata and with programming mastery scoring results.
 The dataset is intended as an input for research on computing education and source code analysis.
+Specifically, it can support assessing the programming skills developed in Scratch, exploring the learning progressions of novice programmers, evaluating the quality of Scratch programs, and identifying smells and bad programming practices.
 \end{abstract}

### 2017-02-17 MSR-Scratch.tex
Updates footnotes with temp dataset repository location

@@
 The dataset contains 250 thousand Scratch projects, from more than 100 thousand different users, that were scraped from the Scratch project repository.
-It is made available as a database\footnote{{\fenia{add anonymized address, update it in repo footnote}}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
+It is made available as a database\footnote{\label{dataseturl}The temporary location of the dataset, for the purpose of the review, is \url{https://github.com/scratchmsrds/scratchmrsds}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
 Scratch quality assessment tool \cite{moreno_automatic_2014}.
@@
 Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
-Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}
+Kragle, as well as all scraped projects and our analysis files are available.\footnotemark[\ref{dataseturl}]
 
@@
 The dataset contains 250 thousand Scratch projects, from more than 100 thousand different users, that were scraped from the Scratch project repository.
-It is made available as a database\footnote{{\fenia{add anonymized address, update it in repo footnote}}} which includes the source code of the Scratch projects, their metadata, and their programming mastery scoring results.
+It is made available as a database\footnotemark[\ref{dataseturl}] which includes the source code of the Scratch projects, their metadata, and their programming mastery scoring results.
 

### 2017-02-17 MSR-Scratch.tex
Removed keywords
Updated abstract

@@
 Scratch is increasingly popular, both as an introductory programming language and as a research target in the computing education research field.	
-In this paper, we present an open dataset of 250K recent Scratch programs from 100K authors scraped from the Scratch project repository.
-We processed the program data to encode it into a database that facilitates querying and further analysis.
-We further enriched the dataset with project metadata and with programming mastery scoring results.
-The dataset is intended as an input for research on computing education and source code analysis.
-Specifically, it can support assessing the programming skills developed in Scratch, exploring the learning progressions of novice programmers, evaluating the quality of Scratch programs, and identifying smells and bad programming practices.
+In this paper, we present a dataset of 250K recent Scratch projects from 100K different authors scraped from the Scratch project repository.
+We processed the projects' source code and metadata to encode them into a database that facilitates querying and further analysis.
+We further evaluated the projects in terms of programming skills and mastery, and we included the project scoring results.
+The dataset enables the analysis of the source code of Scratch projects, of their quality characteristics, and of the programming skills that their authors exhibit.
+The dataset can be used for empirical research in software engineering and computing education.
 \end{abstract}
-
-\begin{IEEEkeywords}
-Scratch; dataset;
-\end{IEEEkeywords}
 

### 2017-02-17 MSR-Scratch.tex
Shortened conclusions

@@
 \section{Conclusion}
-The increasing popularity of Scratch as an introductory programming language makes it an increasingly popular research target in the computing education research field.
-This paper presented an open and timely dataset of recent Scratch programs, that can facilitate research in the fields of source code analysis and computing education.
-The dataset can support various research topics, including the ones related to the assessment of the programming skills that novice programmers develop in the Scratch environment and the exploration of their learning progressions, and issues related to the quality of the programs and the identification of smells and bad programming habits.
-
-The dataset contains 250 thousand Scratch projects, from more than 100 thousand different users, that were scraped from the Scratch project repository.
+We presented a dataset of recent Scratch programs scraped from the Scratch project repository.
 It is made available as a database\footnotemark[\ref{dataseturl}] which includes the source code of the Scratch projects, their metadata, and their programming mastery scoring results.
+The dataset can facilitate research in software engineering and computing education, for topics like the assessment of the programming skills that novice programmers develop, the exploration of their learning progressions, and issues related to the quality of the programs and the identification of smells and bad programming habits.
 

### 2017-02-17 MSR-Scratch.tex
Added limitation
Add Dr. Scratch text and a couple of comments

@@
 \newcommand{\grex}[1]{\emph{\color{yellow}Gregorio says: #1}}
-\newcommand{\jemole}[1]{\emph{\color{red}Jesús says: #1}}
+\newcommand{\jemole}[1]{\emph{\color{red}Jes\'us says: #1}}
 
@@
 A number of works in the computing education research field attempt to assess the programming skills that novice programmers develop in the Scratch environment.
-Some utilize program data collected during specific programming courses (e.g., \cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab, which contains data for Scratch projects created until 2012 (e.g., \cite{fields_2014, yang_2015, Dasgupta_2016}).
+Some utilize program data collected during specific programming courses (e.g., \cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab \jemole{perhaps we could cite the dataset itself~\cite{2017_scratch_dataset}}, which contains data for Scratch projects created until 2012 (e.g., \cite{fields_2014, yang_2015, Dasgupta_2016}).
 In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices within Scratch programs \cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g.
@@
 \subsection{Calculation of programming mastery scores}
-\fenia{Description of how the mastery metrics for       
-	[Abstraction]
-	,[Parallelism]
-	,[Logic]
-	,[Synchronization]
-	,[FlowControl]
-	,[UserInteractivity]
-	,[DataRepresentation]
-	,[Mastery]
-	,[Clones]
-	,[CustomBlocks]
-	,[InstancesSprites]
-	were calculated}
+%\fenia{Description of how the mastery metrics for       
+%	[Abstraction]
+%	,[Parallelism]
+%	,[Logic]
+%	,[Synchronization]
+%	,[FlowControl]
+%	,[UserInteractivity]
+%	,[DataRepresentation]
+%	,[Mastery]
+%	,[Clones]
+%	,[CustomBlocks]
+%	,[InstancesSprites]
+%	were calculated}
+The projects were analyzed with Dr.
+Scratch, a tool that statically inspects Scratch projects source code to assign scores on seven dimensions of computational thinking: abstraction and problem decomposition, logical thinking, synchronization, parallelism, algorithmic notions of flow control, user interactivity and data representation.
+These dimensions are given a value from 0 to 3.
+The mastery score is the aggregated punctuation of these dimensions, and therefore ranges from 0 to 21.
+
+In addition, Dr.
+Scratch also detects intra-project software cloning, i.e.
+repetition of code; the use of custom blocks, which is the way functionality can be reused in Scratch programs; and the use of instances of sprites, a feature labeled in Scratch as clone creation.
+In consequence, these values are also included in the dataset.
 
@@
 		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 project:p\_id}})\\
-		& & \texttt{Abstraction} (\fenia{short descriptions})\\
-		& & \texttt{Parallelism} ()\\
-		& & \texttt{Logic} ()\\
-		& & \texttt{Synchronization} ()\\
-		& & \texttt{FlowControl} ()\\
-		& & \texttt{UserInteractivity} ()\\
-		& & \texttt{DataRepresentation} ()\\
-		& & \texttt{Mastery} ()\\
-		& & \texttt{Clones} ()\\
-		& & \texttt{CustomBlocks} ()\\
-		& & \texttt{InstancesSprites} ()\\
+		& & \texttt{Abstraction} (Dr. Scratch score for abstraction and problem decomposition)\\
+		& & \texttt{Parallelism} (Dr. Scratch score for parallelism)\\
+		& & \texttt{Logic} (Dr. Scratch score for logical thinking)\\
+		& & \texttt{Synchronization} (Dr. Scratch score for synchronization)\\
+		& & \texttt{FlowControl} (Dr. Scratch score for algorithmic notions of flow control)\\
+		& & \texttt{UserInteractivity} (Dr. Scratch score for user interactivity)\\
+		& & \texttt{DataRepresentation} (Dr. Scratch score for data representation)\\
+		& & \texttt{Mastery} (Dr. Scratch total mastery score)\\
+		& & \texttt{Clones} (Number of software clones)\\
+		& & \texttt{CustomBlocks} (Number of user defined blocks)\\
+		& & \texttt{InstancesSprites} (True if project makes use of Scratch clones)\\
 		\hline
@@
 \fenia{Should we add some stats about the Dr Scratch master scores?}
+\jemole{I don't know if we have space for this, but here it goes a proposal: As shown in Table~\ref{tbl-size}, not all the projects were successfully analyzed with Dr.
+Scratch.
+This is due to problems with non-official extension blocks and with some non-ascii characters in sprites names.
+Of those correctly assessed projects, as depicted in Table~\ref{tbl-stats}, the mastery score of the median project is 8, which is considered a medium level of development of computational thinking skills.}
 
@@
 However, we counterbalanced that by collecting a large dataset which comprises around 1.3\% of all 19 million shared Scratch projects, and is still the most recent one to be made available to the research community.
+It must also be noted that the projects in the dataset are the ones that their authors shared publicly when the scraper run.
+The dataset contains cases of projects that are no longer publicly shared and are thus no longer accessible through the Scratch web interface.
 

### 2017-02-17 MSR-Scratch.tex
Beta version

@@
 
-\title{A dataset of Scratch programs,\\Scraped, Shaped and Scored}
+\title{A Dataset of Scratch Programs:\\Scraped, Shaped and Scored}
 
-\author{
-	\IEEEauthorblockN{Efthimia Aivaloglou\IEEEauthorrefmark{1}, Felienne Hermans\IEEEauthorrefmark{1}, Gregorio Robles\IEEEauthorrefmark{2}, and Jes{\'u}s Moreno-Le{\'o}n\IEEEauthorrefmark{3}}
-
-	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
-			Email: \{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
-	\IEEEauthorblockA{\IEEEauthorrefmark{2}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
-		Email: grex@gsyc.urjc.es}
-	\IEEEauthorblockA{\IEEEauthorrefmark{3}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
-		Email: jesus.moreno@programamos.es}
-}
+%\author{
+%	\IEEEauthorblockN{Efthimia Aivaloglou\IEEEauthorrefmark{1}, Felienne Hermans\IEEEauthorrefmark{1}, Jes{\'u}s Moreno-Le{\'o}n\IEEEauthorrefmark{2}, and Gregorio Robles\IEEEauthorrefmark{3}}
+%	
+%	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
+%			Email: \{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
+%	\IEEEauthorblockA{\IEEEauthorrefmark{2}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
+%		Email: jesus.moreno@programamos.es}
+%	\IEEEauthorblockA{\IEEEauthorrefmark{3}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
+%		Email: grex@gsyc.urjc.es}
+%}
 
@@
 We processed the projects' source code and metadata to encode them into a database that facilitates querying and further analysis.
-We further evaluated the projects in terms of programming skills and mastery, and we included the project scoring results.
+We further evaluated the projects in terms of programming skills and mastery, and included the project scoring results.
 The dataset enables the analysis of the source code of Scratch projects, of their quality characteristics, and of the programming skills that their authors exhibit.
@@
 \section{Introduction}
-Scratch \cite{resnick_scratch:_2009} is a block-based programming language developed to serve as a stepping stone for children from 8 to 16 years old to the more advanced world of computer programming.
+Scratch~\cite{resnick_scratch:_2009} is a block-based programming language developed to serve as a stepping stone for children from 8 to 16 years old to the more advanced world of computer programming.
 It offers a web-based programing environment that enables creating games and interactive animations. The public repository of Scratch programs contains over 19 million projects and 16 million users.
@@
 A number of works in the computing education research field attempt to assess the programming skills that novice programmers develop in the Scratch environment.
-Some utilize program data collected during specific programming courses (e.g., \cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab \jemole{perhaps we could cite the dataset itself~\cite{2017_scratch_dataset}}, which contains data for Scratch projects created until 2012 (e.g., \cite{fields_2014, yang_2015, Dasgupta_2016}).
-In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices within Scratch programs \cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g.
-Hairball \cite{boe_hairball:_2013} and Dr.
-Scratch \cite{moreno_automatic_2014}).
+Some utilize program data collected during specific programming courses (e.g.,~\cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab~\cite{2017_scratch_dataset}, which contains data for Scratch projects created until 2012 (e.g.,~\cite{fields_2014, yang_2015, Dasgupta_2016}).
+In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices within Scratch programs~\cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr.
+Scratch~\cite{moreno_automatic_2014}).
 
@@
 The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the Scratch 2 web programming interface in 2013.
-It is since then that the popularity of Scratch started increasing\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
+It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
 
 The goal of this paper is to present an open and timely dataset of recent Scratch programs, along with their metadata, that can facilitate quantitative research in the fields of source code analysis and computing education.
-The dataset contains 250 thousand Scratch projects, from more than 100 thousand different users, that were scraped from the Scratch project repository.
+The dataset contains 250,000 Scratch projects, from more than 100,000 different users, that were scraped from the Scratch project repository.
 It is made available as a database\footnote{\label{dataseturl}The temporary location of the dataset, for the purpose of the review, is \url{https://github.com/scratchmsrds/scratchmrsds}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
-Scratch quality assessment tool \cite{moreno_automatic_2014}.
+Scratch quality assessment tool~\cite{moreno_automatic_2014}.
 
@@
 \subsection{Data collection}
-To collect the data from the web interface of the Scratch project repository we built a scraping program.
+To collect the data from the web interface of the Scratch project repository, we built a scraping program.
 The web scraping program, called Kragle, starts by reading the Scratch projects page\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} and thus obtains the project identifiers of projects that were most recently shared.
-Subsequently, Kragle retrieves the JSON code for each of the listed projects\footnote{For a given project id $x$, the program's JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}}.
+Subsequently, Kragle retrieves a JSON file for each of the listed projects\footnote{For a given project id $x$, the program's JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}}.
 
-We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON code for \nPrograms projects.
+We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects.
 Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
-Kragle, as well as all scraped projects and our analysis files are available.\footnotemark[\ref{dataseturl}]
+Kragle, all scraped projects and our analysis files are available\footnotemark[\ref{dataseturl}].
 
@@
 For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
-We included the blocks from Scratch extensions, such as the LEGO WeDo extensions, that we found in the dataset.
+We included blocks from Scratch extensions, such as the LEGO WeDo extensions, that were found in the dataset.
 
@@
 
-In addition, Dr.
-Scratch also detects intra-project software cloning, i.e.
-repetition of code; the use of custom blocks, which is the way functionality can be reused in Scratch programs; and the use of instances of sprites, a feature labeled in Scratch as clone creation.
-In consequence, these values are also included in the dataset.
+Dr.
+Scratch also detects: intra-project software cloning, i.e., repetition of code; the use of custom blocks, which is the way functionality can be reused in Scratch programs; and the use of instances of sprites, a feature labeled in Scratch as clone creation.
+These values are included in the dataset.
 
@@
 We also imported the data on the shapes and the categories of the Scratch blocks.
-We then used SQL queries for normalizing the data and bringing it in its final schema, outlined in Table \ref{tbl-dbschema}.
+We then used SQL queries for normalizing the data and bringing it in its final schema, outlined in Table~\ref{tbl-dbschema}.
 
@@
 		&  & 							\texttt{total-favorites} (total users favoriting)\\
-		&  & 							\texttt{total-loves} (total users `loving' the project)\\
+		&  & 							\texttt{total-loves} (users `loving' the project)\\
 		& & 							\texttt{is-remix} (calculated column, true if project is a remix of another one)\\
@@
 		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 project:p\_id}})\\
-		& & 							\texttt{sprite-type} (one of [sprite, stage, procDef])\\
+		& & 							\texttt{sprite-type} ([$sprite \vert stage \vert procDef$])\\
 		& & 							\texttt{sprite-name} (name sprite the script is in)\\
@@
 		&  & \texttt{parameter-1} (value of 1st input parameter)\\
-		&  &\vdots\\
-		&  & \texttt{parameter-24} (value of 24th input parameter)\\
+		&  & ... \\
+		&  & \texttt{parameter-24} (value 24th input parameter)\\
 		\hline
@@
 		&  & \texttt{category} (Scratch block category)\\
-		&  & \texttt{shape} (One of the 8 block shapes in the Scratch interface)\\
-		&  & \texttt{is-input} (true if the block receives user input)\\
+		&  & \texttt{shape} (One of 8 block shapes in Scratch)\\
+		&  & \texttt{is-input} (true if block receives user input)\\
 		\hline
@@
 		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 project:p\_id}})\\
-		& & \texttt{Abstraction} (Dr. Scratch score for abstraction and problem decomposition)\\
-		& & \texttt{Parallelism} (Dr. Scratch score for parallelism)\\
-		& & \texttt{Logic} (Dr. Scratch score for logical thinking)\\
-		& & \texttt{Synchronization} (Dr. Scratch score for synchronization)\\
-		& & \texttt{FlowControl} (Dr. Scratch score for algorithmic notions of flow control)\\
-		& & \texttt{UserInteractivity} (Dr. Scratch score for user interactivity)\\
-		& & \texttt{DataRepresentation} (Dr. Scratch score for data representation)\\
+% Original
+%		& & \texttt{Abstraction} (Score for abstraction and problem decomposition)\\
+%		& & \texttt{Parallelism} (Score for parallelism)\\
+%		& & \texttt{Logic} (Score for logical thinking)\\
+%		& & \texttt{Synchronization} (Score for synchronization)\\
+%		& & \texttt{FlowControl} (Score for algorithmic notions of flow control)\\
+%		& & \texttt{UserInteractivity} (Score for user interactivity)\\
+%		& & \texttt{DataRepresentation} (Score for data representation)\\
+
+		& & \texttt{Abstraction} (Score)\\
+		& & \texttt{Parallelism} (Score)\\
+		& & \texttt{Logic} (Score)\\
+		& & \texttt{Synchronization} (Score)\\
+		& & \texttt{FlowControl} (Score)\\
+		& & \texttt{UserInteractivity} (Score)\\
+		& & \texttt{DataRepresentation} (Score)\\
 		& & \texttt{Mastery} (Dr. Scratch total mastery score)\\
-		& & \texttt{Clones} (Number of software clones)\\
-		& & \texttt{CustomBlocks} (Number of user defined blocks)\\
-		& & \texttt{InstancesSprites} (True if project makes use of Scratch clones)\\
+		& & \texttt{Clones} (\# software clones)\\
+		& & \texttt{CustomBlocks} (\# user defined blocks)\\
+		& & \texttt{InstancesSprites} (True if project uses Scratch clones)\\
 		\hline
@@
 
-The projects' data are stored in a relational database.
+The data of a project are stored in a relational database.
 Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{\label{fn-authorpage}Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
 If a project is a remix of another one, the original project can be found in the \textbf{\texttt{remix}} table.
-Table \textbf{\texttt{grades}} stores the Dr. Scratch results for the programming mastery metrics per project.
+Table~\textbf{\texttt{grades}} stores the Dr. Scratch results for the programming mastery metrics per project.
 
@@
 
-The schema of the database, outlined in Table \ref{tbl-dbschema}, reflects the structure of the Scratch programs.
+The schema of the database, outlined in Table~\ref{tbl-dbschema}, reflects the structure of the Scratch programs.
 Projects contain sprites, which are entities with their own associated code.
 The code is organized into \textbf{\texttt{script}}s, i.e., groups of Scratch code blocks, each script belonging a sprite named \texttt{sprite-name}.
-In the example in Figure \ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
+In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
 Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedure}} table.
@@
 
-\begin{table*}[]
+\begin{table}[]
 	\centering
@@
 		\hline
-		Projects per username&2.27&1&1&1&2&868\\
-		Sprites per project&5.30&0&1&2&5&525\\
-		Scripts per project&16.19&0&2&4&11&3,038\\
-		Blocks per project&144.25&0&9&26&69&34,622\\
-		Custom block definitions per project&0.64&0&0&0&0&372\\
-		Total views per project&5.76&0&1&1&4&27,993\\
-		Total remixes per project&197.12&0&0&0&0&14,951,164\\
-		Total favorites per project&0.54&0&0&0&0&2,582\\
-		Dr. Scratch mastery score per project&8.92&0&6&8&11&21\\
+		Projects per username&2.3&1&1&1&2&868\\
+		Sprites \emph{pp}&5.3&0&1&2&5&525\\
+		Scripts \emph{pp}&16.2&0&2&4&11&3,038\\
+		Blocks \emph{pp}&144.3&0&9&26&69&34,622\\
+		Custom block def. \emph{pp}&0.6&0&0&0&0&372\\
+		Total views \emph{pp}&5.8&0&1&1&4&27,993\\
+		Total remixes \emph{pp}&197.1&0&0&0&0&14,951K \\ %14,951,164  \\
+		Total favorites \emph{pp}&0.5&0&0&0&0&2,582\\
+		Dr.Scratch mastery \emph{pp}&8.9&0&6&8&11&21\\
 		\hline
 	\end{tabular}
-	\caption{Summary statistics of the dataset}
+	\caption{Summary statistics of the dataset (\emph{pp} stands for per project).}
 	\label{tbl-stats}
-\end{table*}
+\end{table}
+
+% Original table
+%\begin{table*}[]
+%	\centering
+%	\begin{tabular}{lrrrrrr}
+%		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
+%		\hline
+%		Projects per username&2.27&1&1&1&2&868\\
+%		Sprites per project&5.30&0&1&2&5&525\\
+%		Scripts per project&16.19&0&2&4&11&3,038\\
+%		Blocks per project&144.25&0&9&26&69&34,622\\
+%		Custom block definitions per project&0.64&0&0&0&0&372\\
+%		Total views per project&5.76&0&1&1&4&27,993\\
+%		Total remixes per project&197.12&0&0&0&0&14,951,164\\
+%		Total favorites per project&0.54&0&0&0&0&2,582\\
+%		Dr. Scratch mastery score per project&8.92&0&6&8&11&21\\
+%		\hline
+%	\end{tabular}
+%	\caption{Summary statistics of the dataset}
+%	\label{tbl-stats}
+%\end{table*}
 
@@
 		Non-empty projects & 233,491 \\
+		Projects analyzed by Dr. Scratch & 231,050 \\
 		Remixes of other projects & 12,167 \\
@@
 		Custom block definitions & 209,769 \\
-		Projects analyzed by Dr. Scratch & 231,050 \\
 		\hline
@@
 
-As shown in Table \ref{tbl-size}, the dataset contains the metadata of 250,163 Scratch projects and the code of 233,491 of those that are non-empty.
-The projects were created from 109,960 different users.
-Most of those, as shown in Table \ref{tbl-stats}, have created a single project.
-However, 40 thousand users have created more than one and up to 868 projects in the dataset.
+As shown in Table \ref{tbl-size}, the dataset contains the metadata of 250,163 Scratch projects, and the code of 233,491 of those that are non-empty.
+The projects were created by 109,960 different users.
+Most of the users, as shown in Table~\ref{tbl-stats}, have created a single project.
+However, 40,000 users have created more than one and up to 868 projects in the dataset.
+Not all the projects were successfully analyzed with Dr. Scratch. This is due to problems with non-official extension blocks and with some non-ascii characters in sprites names.
 
 The majority of the projects have been viewed only once, and they have not been favorited or remixed.
-Around 10 thousand of the total projects have been remixed at least once, while the dataset includes cases of very popular projects with more than 100 remixes and even one thousand cases with more than 100 views.
+Around 10,000 of the total projects have been remixed at least once, while the dataset includes cases of very popular projects with more than 100 remixes and even one thousand cases with more than 100 views.
 
-Table \ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21.
-The table also highlights the existence of surprisingly complex projects, with 525 sprites, 3 thousand scripts as 34 thousand blocks.
-
-\fenia{Should we add some stats about the Dr Scratch master scores?}
-\jemole{I don't know if we have space for this, but here it goes a proposal: As shown in Table~\ref{tbl-size}, not all the projects were successfully analyzed with Dr.
-Scratch.
-This is due to problems with non-official extension blocks and with some non-ascii characters in sprites names.
-Of those correctly assessed projects, as depicted in Table~\ref{tbl-stats}, the mastery score of the median project is 8, which is considered a medium level of development of computational thinking skills.}
+Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21, which is considered a medium level of development of computational thinking skills.
+The table highlights the existence of surprisingly complex projects, with 525 sprites, 3,000 scripts and 34,000 blocks.
 
 \section{Enabled Research and Dataset Extension}
-\fenia{What about our own papers utilizing the dataset?
+
+%\fenia{What about our own papers utilizing the dataset?
 (\cite{Aivaloglou_2016} and \cite{Robles_2017}).
@@
 The dataset can be used for examining indications of learning of programming concepts and abstractions.
-Existing works in this research direction include the work by Maloney \emph{et al.} \cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers, and found that within the least utilized ones are boolean operators and variables.
-Another study on the internalization of programming concepts with Scratch with 46 students was presented in \cite{meerbaum-salant_learning_2010}.
+Existing works in this research direction include the work by Maloney \emph{et al.}~\cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers, and found that within the least utilized ones are boolean operators and variables.
+Another study on the internalization of programming concepts with Scratch with 46 students was presented in~\cite{meerbaum-salant_learning_2010}.
 Concepts like loops, conditional loops, message passing, initialization, variables and concurency were examined, and it was found that students had problems with the last three.
-In \cite{wilson_evaluation_2012}, 29 projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
+In~\cite{wilson_evaluation_2012}, 29 projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
 
@@
 Another promising research direction relates to the \textit{quality of the programming artifacts developed in the Scratch environment}.
-The dataset can be used for quality assessment, for the identification of programming smells, and for exploring indications of harmful programming habits.
-Existing works in this direction include a study in a classroom setting \cite{Meerbaum_habits_2011}, which highlighted two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming.
+The dataset can be used for quality assessment, for the identification of programming smells~\cite{Aivaloglou_2016}, and for exploring indications of harmful programming habits~\cite{Robles_2017}.
+Existing works in this direction include a study in a classroom setting~\cite{Meerbaum_habits_2011}, which highlighted two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming.
 The authors connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops.
-Related to smell detection are the Scratch automated quality analysis tools Hairball \cite{boe_hairball:_2013}, a lint-like static analysis tool for Scratch that can detect initialization problems and unmatched broadcast and receive blocks, and Dr.
-Scratch \cite{moreno-leon_dr._2015}, which extends Hairball to detect two bad programming habits in Scratch: not changing the default object names and duplicating scripts.
+Related to smell detection are the Scratch automated quality analysis tools Hairball~\cite{boe_hairball:_2013}, a lint-like static analysis tool for Scratch that can detect initialization problems and unmatched broadcast and receive blocks, and Dr.
+Scratch~\cite{moreno-leon_dr._2015}, which extends Hairball to detect two bad programming habits in Scratch: not changing the default object names and duplicating scripts.
 
 This dataset will enable examining code quality and code smells on a large set of Scratch projects.
-Static analysis of the source code can also be performed for the identification of other types of smells \cite{fowler_refactoring:_1999} that might be common in the artifacts of novice programmers.
+Static analysis of the source code can also be performed for the identification of other types of smells~\cite{fowler_refactoring:_1999} that might be common in the artifacts of novice programmers.
 
@@
 Existing works in this area have used the dataset of Scratch programs created until 2012, before the current web-based programming environment was introduced, and described in the Introduction.
-They include the works of Dasgupta et al., who investigated how project remixing relates to the adoption of new computational thinking concepts \cite{Dasgupta_2016}, and Yang et al., who examined the learning patterns of programmers in terms of block use over their first 50 projects \cite{yang_2015}.
+They include the works of Dasgupta et al., who investigated how project remixing relates to the adoption of new computational thinking concepts~\cite{Dasgupta_2016}, and Yang et al., who examined the learning patterns of programmers in terms of block use over their first 50 projects~\cite{yang_2015}.
 
@@
 Apart from quantitative studies, the dataset can support the design of experiments and field studies, whose material often includes Scratch programs with specific characteristics.
-For example, in \cite{hermans_2016} examples of Scratch programs exhibiting specific smells were used in a controlled experiment to determine how the smells affected the understanding and the modification of the programs.
+For example, in~\cite{hermans_2016} examples of Scratch programs exhibiting specific smells were used in a controlled experiment to determine how the smells affected the understanding and the modification of the programs.
 This queryable dataset can support finding example programs with characteristics according to the experiment design and the field of the study.
@@
 This suffices for creating project portfolios of the authors and facilitating research in the directions described in the previous section.
-However, other directions in the computing education research field require richer author data, and especially data on their gender and their age.
-There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman \cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelization, conditionals and, especially, variables were under-represented by all grades apart from 5 an 6.
-The effect of gender and account age were also examined in \cite{fields_2014} in relation to the use of programming concepts.
-Studies on the effect of parameters like the age are not only interesting but also of direct value to educators, since they give insights on when it is better to introduce computational thinking concepts in the curriculum of schools.
-However, this dataset does not include information on the gender and the age of authors and cannot be extended to do so using the current Scratch web interface because this information is not available in the user profile page\footnotemark[\ref{fn-authorpage}], even though it is provided by the users upon registration.
+However, other directions in the computing education research field require richer author data, and especially data on their gender and age.
+There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman~\cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelization, conditionals and, especially, variables were under-represented by all grades apart from 5 an 6.
+The effect of gender and account age were also examined in~\cite{fields_2014} in relation to the use of programming concepts.
+%Studies on the effect of parameters like the age are not only interesting but also of direct value to educators, since they give insights on when it is better to introduce computational thinking concepts in the curriculum of schools.
+However, this dataset does not include information on the gender and the age of authors and cannot be extended to do so using the current Scratch web interface because this information is not available in the user profile page\footnotemark[\ref{fn-authorpage}], even though it is provided by users upon registration.
 
@@
 It could be the case that the programming habits of Scratch users are changing over time.
-However, we counterbalanced that by collecting a large dataset which comprises around 1.3\% of all 19 million shared Scratch projects, and is still the most recent one to be made available to the research community.
-It must also be noted that the projects in the dataset are the ones that their authors shared publicly when the scraper run.
+However, we counterbalanced that by collecting a large dataset which comprises around 1.3\% of all 19 million shared Scratch projects, and is the most recent one to be made available to the research community.
+Projects in the dataset are the ones that their authors shared publicly when the scraper run.
 The dataset contains cases of projects that are no longer publicly shared and are thus no longer accessible through the Scratch web interface.
@@
 It is made available as a database\footnotemark[\ref{dataseturl}] which includes the source code of the Scratch projects, their metadata, and their programming mastery scoring results.
-The dataset can facilitate research in software engineering and computing education, for topics like the assessment of the programming skills that novice programmers develop, the exploration of their learning progressions, and issues related to the quality of the programs and the identification of smells and bad programming habits.
+The dataset can facilitate research in software engineering and computing education, for topics like the assessment of the programming skills that novice programmers develop, the exploration of their learning progressions, and issues related to the quality of the programs, such as the identification of smells and bad programming habits.
 
 \bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,ScratchDataset}
+\bibliography{ScratchDataset}
 

### 2017-02-17 MSR-Scratch.tex
Beta2 version

@@
 Scratch~\cite{resnick_scratch:_2009} is a block-based programming language developed to serve as a stepping stone for children from 8 to 16 years old to the more advanced world of computer programming.
-It offers a web-based programing environment that enables creating games and interactive animations. The public repository of Scratch programs contains over 19 million projects and 16 million users.
+It offers a web-based programming environment that enables creating games and interactive animations. The public repository of Scratch programs contains over 19 million projects and 16 million users.
 
@@
 These dimensions are given a value from 0 to 3.
-The mastery score is the aggregated punctuation of these dimensions, and therefore ranges from 0 to 21.
+The mastery score is the aggregated punctuation of the seven dimensions, and therefore ranges from 0 to 21.
 
@@
 However, 40,000 users have created more than one and up to 868 projects in the dataset.
-Not all the projects were successfully analyzed with Dr. Scratch. This is due to problems with non-official extension blocks and with some non-ascii characters in sprites names.
+Not all the projects were successfully analyzed with Dr. Scratch. This is due to problems with non-official extension blocks and with some non-ASCII characters in sprites names.
 
@@
 The dataset can be used for examining indications of learning of programming concepts and abstractions.
-Existing works in this research direction include the work by Maloney \emph{et al.}~\cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers, and found that within the least utilized ones are boolean operators and variables.
+Existing works in this research direction include the work by Maloney \emph{et al.}~\cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers, and found that within the least utilized ones are Boolean operators and variables.
 Another study on the internalization of programming concepts with Scratch with 46 students was presented in~\cite{meerbaum-salant_learning_2010}.
-Concepts like loops, conditional loops, message passing, initialization, variables and concurency were examined, and it was found that students had problems with the last three.
+Concepts like loops, conditional loops, message passing, initialization, variables and concurrency were examined, and it was found that students had problems with the last three.
 In~\cite{wilson_evaluation_2012}, 29 projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
@@
 However, other directions in the computing education research field require richer author data, and especially data on their gender and age.
-There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman~\cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelization, conditionals and, especially, variables were under-represented by all grades apart from 5 an 6.
+There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman~\cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelism, conditionals and, especially, variables were under-represented by all grades apart from 5 an 6.
 The effect of gender and account age were also examined in~\cite{fields_2014} in relation to the use of programming concepts.

### 2017-02-19 MSR-Scratch.tex
Fixed error on remix stats

@@
 		Total views \emph{pp}&5.8&0&1&1&4&27,993\\
-		Total remixes \emph{pp}&197.1&0&0&0&0&14,951K \\ %14,951,164  \\
+		Total remixes \emph{pp}&0.1&0&0&0&0&306\\
 		Total favorites \emph{pp}&0.5&0&0&0&0&2,582\\

### 2017-02-20 MSR-Scratch.tex
Small fixes

@@
 Some utilize program data collected during specific programming courses (e.g.,~\cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab~\cite{2017_scratch_dataset}, which contains data for Scratch projects created until 2012 (e.g.,~\cite{fields_2014, yang_2015, Dasgupta_2016}).
-In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices within Scratch programs~\cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr.
+In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices ~\cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr.
 Scratch~\cite{moreno_automatic_2014}).
@@
 While Scratch is receiving increasing interest as an introductory programming language, there is no recent dataset of Scratch programs available to the research community.
-The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the Scratch 2 web programming interface in 2013.
+The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the current Scratch version (Scratch 2) and of the web programming interface in 2013.
 It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
@@
 To collect the data from the web interface of the Scratch project repository, we built a scraping program.
-The web scraping program, called Kragle, starts by reading the Scratch projects page\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} and thus obtains the project identifiers of projects that were most recently shared.
-Subsequently, Kragle retrieves a JSON file for each of the listed projects\footnote{For a given project id $x$, the program's JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}}.
+The web scraping program starts by reading the Scratch projects page\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} and thus obtains the project identifiers of projects that were most recently shared.
+Subsequently, it retrieves a JSON file for each of the listed projects\footnote{For a given project id $x$, the program's JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}}.
 
-We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects.
+We ran the scraper on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects.
 Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
-Kragle, all scraped projects and our analysis files are available\footnotemark[\ref{dataseturl}].
+The scraper and all project files are available\footnotemark[\ref{dataseturl}].
 
@@
 This resulted in a list of used blocks per project, within the sprites and the stage of the project.
-We also cross referenced all blocks with the Scratch wiki to determine the shapes and the category of all blocks.
+We also cross referenced all blocks with the Scratch wiki to determine their shapes and category.
 For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
@@
 %	were calculated}
-The projects were analyzed with Dr.
-Scratch, a tool that statically inspects Scratch projects source code to assign scores on seven dimensions of computational thinking: abstraction and problem decomposition, logical thinking, synchronization, parallelism, algorithmic notions of flow control, user interactivity and data representation.
+We analyzed the projects with Dr.
+Scratch, a tool that statically inspects Scratch projects' source code to assign scores on seven dimensions of computational thinking: abstraction and problem decomposition, logical thinking, synchronization, parallelism, algorithmic notions of flow control, user interactivity and data representation.
 These dimensions are given a value from 0 to 3.
@@
 These values are included in the dataset.
+
+Out of the \nPrograms~projects, 231,050 were successfully analyzed with Dr. Scratch. This was due to problems with non-official extension blocks and with some non-ASCII characters in sprites names.
 
@@
 		& & 							\texttt{sprite-type} ([$sprite \vert stage \vert procDef$])\\
-		& & 							\texttt{sprite-name} (name sprite the script is in)\\
+		& & 							\texttt{sprite-name} (name of sprite the script is in)\\
 		& & 							\texttt{script-rank} (project-level ranking of script)\\
@@
 		& PK & \texttt{block-rank} (script-level ranking of block)\\
-		& FK & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockType:b-type}})\\
+		&  & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockType:b-type}})\\
 		&  & \texttt{parameter-1} (value of 1st input parameter)\\
@@
 		& & \texttt{Mastery} (Dr. Scratch total mastery score)\\
-		& & \texttt{Clones} (\# software clones)\\
-		& & \texttt{CustomBlocks} (\# user defined blocks)\\
-		& & \texttt{InstancesSprites} (True if project uses Scratch clones)\\
+		& & \texttt{Clones} (number of software clones)\\
+		& & \texttt{CustomBlocks} (number of user defined blocks)\\
+		& & \texttt{InstancesSprites} (true if project uses Scratch clones)\\
 		\hline
@@
 
-The data of a project are stored in a relational database.
+The projects' data are stored in a relational database.
 Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{\label{fn-authorpage}Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
@@
 Table~\textbf{\texttt{grades}} stores the Dr. Scratch results for the programming mastery metrics per project.
+
+The schema of the database, outlined in Table~\ref{tbl-dbschema}, reflects the structure of the Scratch programs.
+Projects contain sprites, which are entities with their own associated code.
+The code is organized into \textbf{\texttt{script}}s, i.e., groups of Scratch code blocks, each script belonging to a sprite named \texttt{sprite-name}.
+In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
+Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedure}} table.
 
@@
 
-The schema of the database, outlined in Table~\ref{tbl-dbschema}, reflects the structure of the Scratch programs.
-Projects contain sprites, which are entities with their own associated code.
-The code is organized into \textbf{\texttt{script}}s, i.e., groups of Scratch code blocks, each script belonging a sprite named \texttt{sprite-name}.
-In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
-Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedure}} table.
-
 The coding elements in the Scratch environment are the \textbf{\texttt{block}}s, which can receive input parameters.
@@
 The clocks can be of different shapes and categories.
-In table \textbf{\texttt{blockType}} we have stored the encoding of all types of Scratch blocks and the blocks from Scratch extensions.
+In table \textbf{\texttt{blockType}} we have stored the encoding of all types of Scratch blocks and of the blocks from Scratch extensions.
 
@@
 		Blocks & 36,085,654 \\
-		Custom block definitions & 209,769 \\
+		Custom block definitions & 204,018 \\
 		\hline
@@
 However, 40,000 users have created more than one and up to 868 projects in the dataset.
-Not all the projects were successfully analyzed with Dr. Scratch. This is due to problems with non-official extension blocks and with some non-ASCII characters in sprites names.
 

### 2017-02-20 MSR-Scratch.tex
More fixes

@@
 It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
+Scratch 2 introduced several features relating to important programming concepts,\footnote{The complete list of features introduced in Scratch 2 can be found at \url{https://wiki.scratch.mit.edu/wiki/Scratch\_2.0}} like custom blocks (the equivalent to procedure definitions), and therefore that dataset does not include projects utilizing those features.
 
@@
 The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education is receiving significant research attention.
-A number of studies have been carried out during the last years on understanding of programming concepts and the programming practices of novice programmers in block-based environments, on the programming skills they develop, and on the quality of Scratch programs.
+A number of studies have been carried out during the last years on understanding of the programming concepts and the programming practices of novice programmers in those environments, on the programming skills they develop, and on the quality of their programs.
 
@@
 The dataset can be used for examining indications of learning of programming concepts and abstractions.
-Existing works in this research direction include the work by Maloney \emph{et al.}~\cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers, and found that within the least utilized ones are Boolean operators and variables.
-Another study on the internalization of programming concepts with Scratch with 46 students was presented in~\cite{meerbaum-salant_learning_2010}.
-Concepts like loops, conditional loops, message passing, initialization, variables and concurrency were examined, and it was found that students had problems with the last three.
-In~\cite{wilson_evaluation_2012}, 29 projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
+Existing works in this research direction include the work by Maloney \emph{et al.}~\cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts and found that within the least utilized ones are Boolean operators and variables.
+Another study on the internalization of programming concepts with Scratch with 46 students was presented in~\cite{meerbaum-salant_learning_2010}, where it was found that students had problems with initialization, variables and concurrency.
 
@@
 This includes examining the factors that support learning and improving their programming skills.
-Existing works in this area have used the dataset of Scratch programs created until 2012, before the current web-based programming environment was introduced, and described in the Introduction.
+Existing works in this area have used the dataset of the previous version of Scratch programs, created until 2012, and described in the Introduction.
 They include the works of Dasgupta et al., who investigated how project remixing relates to the adoption of new computational thinking concepts~\cite{Dasgupta_2016}, and Yang et al., who examined the learning patterns of programmers in terms of block use over their first 50 projects~\cite{yang_2015}.
@@
 Apart from quantitative studies, the dataset can support the design of experiments and field studies, whose material often includes Scratch programs with specific characteristics.
-For example, in~\cite{hermans_2016} examples of Scratch programs exhibiting specific smells were used in a controlled experiment to determine how the smells affected the understanding and the modification of the programs.
+For example, in~\cite{hermans_2016} examples of Scratch programs exhibiting specific smells were used in a controlled experiment to determine how the smells affect the understanding and the modification of the programs.
 This queryable dataset can support finding example programs with characteristics according to the experiment design and the field of the study.
@@
 The only information about the authors of the Scratch projects contained in the dataset is their username.
-This suffices for creating project portfolios of the authors and facilitating research in the directions described in the previous section.
+This suffices for creating project portfolios of the authors and for facilitating research in the directions described in the previous section.
 However, other directions in the computing education research field require richer author data, and especially data on their gender and age.
@@
 
-Another limitation of this dataset is that we did not scrape a random sample of Scratch projects, but the 250,000 most recent ones during the time that we run the scraper.
+Another limitation of this dataset is that we did not scrape a random sample of Scratch projects, but the most recent ones during the time that we run the scraper.
 It could be the case that the programming habits of Scratch users are changing over time.

### 2017-02-20 MSR-Scratch.tex
Some more minor fixes

@@
 The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the current Scratch version (Scratch 2) and of the web programming interface in 2013.
-It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
-Scratch 2 introduced several features relating to important programming concepts,\footnote{The complete list of features introduced in Scratch 2 can be found at \url{https://wiki.scratch.mit.edu/wiki/Scratch\_2.0}} like custom blocks (the equivalent to procedure definitions), and therefore that dataset does not include projects utilizing those features.
+It is since then that the popularity of Scratch started to increase.\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}
+Scratch 2 introduced several features relating to important programming concepts,\footnote{The complete list of features introduced in Scratch 2 can be found at \url{https://wiki.scratch.mit.edu/wiki/Scratch_2.0}} like custom blocks (the equivalent to procedure definitions), and therefore that dataset does not include projects utilizing those features.
 
@@
 
-\subsection{Data collection}
+\subsection{Data Collection}
 To collect the data from the web interface of the Scratch project repository, we built a scraping program.
@@
 
-\subsection{Calculation of programming mastery scores}
+\subsection{Calculation of Programming Mastery Scores}
 %\fenia{Description of how the mastery metrics for       
@@
 
-\subsection{Importing the data}
+\subsection{Importing the Data}
 \label{dataAnalysis}

### 2017-03-14 MSR-Scratch.tex
Updated DB table names (convert to plural)

@@
 		\hline
-		\textbf{\texttt{project}} & PK & \texttt{p\_id} (Scratch project ID)\\
+		\textbf{\texttt{projects}} & PK & \texttt{p\_id} (Scratch project ID)\\
 		&  & 							\texttt{project-name} (name given to project)\\
@@
 		\hline
-		\textbf{\texttt{script}} & PK & \texttt{script\_id} (auto increment)\\
-		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 project:p\_id}})\\
+		\textbf{\texttt{scripts}} & PK & \texttt{script\_id} (auto increment)\\
+		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 projects:p\_id}})\\
 		& & 							\texttt{sprite-type} ([$sprite \vert stage \vert procDef$])\\
@@
 		\hline
-		\textbf{\texttt{procedure}} & PK,FK & \texttt{script\_id} (\texttt{\scriptsize{1\ldots1 script:script\_id}})\\
+		\textbf{\texttt{procedures}} & PK,FK & \texttt{script\_id} (\texttt{\scriptsize{1\ldots1 scripts:script\_id}})\\
 		& & 							\texttt{proc-name} (name given to custom block)\\
@@
 		\hline
-		\textbf{\texttt{block}} & PK,FK  & \texttt{script\_id} (\texttt{\scriptsize{*\ldots1 script:script\_id}})\\
+		\textbf{\texttt{blocks}} & PK,FK  & \texttt{script\_id} (\texttt{\scriptsize{*\ldots1 scripts:script\_id}})\\
 		& PK & \texttt{block-rank} (script-level ranking of block)\\
-		&  & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockType:b-type}})\\
+		&  & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockTypes:b-type}})\\
 		&  & \texttt{parameter-1} (value of 1st input parameter)\\
@@
 		\hline
-		\textbf{\texttt{blockType}} & PK & \texttt{b-type} (Scratch name of predefined block)\\
+		\textbf{\texttt{blockTypes}} & PK & \texttt{b-type} (Scratch name of predefined block)\\
 		&  & \texttt{category} (Scratch block category)\\
@@
 		\hline
-		\textbf{\texttt{remix}} & PK  & \texttt{remix\_id} (\texttt{\scriptsize{project\_id}} of created project)\\
+		\textbf{\texttt{remixes}} & PK  & \texttt{remix\_id} (\texttt{\scriptsize{project\_id}} of created project)\\
 		&  & \texttt{from-project\_id} (\texttt{\scriptsize{project\_id}} of original project)\\
 		\hline
-		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 project:p\_id}})\\
+		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 projects:p\_id}})\\
 % Original
@@
 The projects' data are stored in a relational database.
-Each \textbf{\texttt{project}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{\label{fn-authorpage}Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
-If a project is a remix of another one, the original project can be found in the \textbf{\texttt{remix}} table.
+Each of the \textbf{\texttt{projects}} is identified by its Scratch project ID, stored in field \texttt{p\_id}, while its author is identified by the \texttt{username}\footnote{\label{fn-authorpage}Each project and author page can be accessed in the Scratch web interface at \url{https://scratch.mit.edu/projects/<p_id>} and \url{https://scratch.mit.edu/users/<username>} respectively}.
+If a project is a remix of another one, the original project can be found in the \textbf{\texttt{remixes}} table.
 Table~\textbf{\texttt{grades}} stores the Dr. Scratch results for the programming mastery metrics per project.
@@
 Projects contain sprites, which are entities with their own associated code.
-The code is organized into \textbf{\texttt{script}}s, i.e., groups of Scratch code blocks, each script belonging to a sprite named \texttt{sprite-name}.
+The code is organized into \textbf{\texttt{scripts}}, i.e., groups of Scratch code blocks, each script belonging to a sprite named \texttt{sprite-name}.
 In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
-Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedure}} table.
+Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedures}} table.
 
@@
 
-The coding elements in the Scratch environment are the \textbf{\texttt{block}}s, which can receive input parameters.
+The coding elements in the Scratch environment are the \textbf{\texttt{blocks}}, which can receive input parameters.
 In the example in the figure, block `go to x:0 y:0' receives as input two numerical parameters.
@@
 The clocks can be of different shapes and categories.
-In table \textbf{\texttt{blockType}} we have stored the encoding of all types of Scratch blocks and of the blocks from Scratch extensions.
+In table \textbf{\texttt{blockTypes}} we have stored the encoding of all types of Scratch blocks and of the blocks from Scratch extensions.
 

### 2017-03-14 MSR-Scratch.tex
De-anonymized

@@
 
-%\author{
-%	\IEEEauthorblockN{Efthimia Aivaloglou\IEEEauthorrefmark{1}, Felienne Hermans\IEEEauthorrefmark{1}, Jes{\'u}s Moreno-Le{\'o}n\IEEEauthorrefmark{2}, and Gregorio Robles\IEEEauthorrefmark{3}}
-%	
-%	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
-%			Email: \{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
-%	\IEEEauthorblockA{\IEEEauthorrefmark{2}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
-%		Email: jesus.moreno@programamos.es}
-%	\IEEEauthorblockA{\IEEEauthorrefmark{3}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
-%		Email: grex@gsyc.urjc.es}
-%}
+\author{
+	\IEEEauthorblockN{Efthimia Aivaloglou\IEEEauthorrefmark{1}, Felienne Hermans\IEEEauthorrefmark{1}, Jes{\'u}s Moreno-Le{\'o}n\IEEEauthorrefmark{2}, and Gregorio Robles\IEEEauthorrefmark{3}}
+	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
+			Email: \{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
+	\IEEEauthorblockA{\IEEEauthorrefmark{2}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
+		Email: jesus.moreno@programamos.es}
+	\IEEEauthorblockA{\IEEEauthorrefmark{3}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
+		Email: grex@gsyc.urjc.es}
+}
 
@@
 The dataset contains 250,000 Scratch projects, from more than 100,000 different users, that were scraped from the Scratch project repository.
-It is made available as a database\footnote{\label{dataseturl}The temporary location of the dataset, for the purpose of the review, is \url{https://github.com/scratchmsrds/scratchmrsds}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
+It is made available as a database\footnote{\label{dataseturl}\url{https://github.com/TUDelftScratchLab/ScratchDataset}} which includes, for each Scratch project, its metadata and the program data, along with programming mastery scoring results from the Dr.
 Scratch quality assessment tool~\cite{moreno_automatic_2014}.
@@
 
-%\fenia{What about our own papers utilizing the dataset?
-(\cite{Aivaloglou_2016} and \cite{Robles_2017}).
-We have to anonymize the submission, so do we add them now or add a note that we omitted them and add them to the camera ready?}
-
 The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education is receiving significant research attention.
@@
 The first research direction that this dataset can be utilized for concerns the \textit{assessment of the programming skills that novice programmers develop in the Scratch environment}.
-The dataset can be used for examining indications of learning of programming concepts and abstractions.
+The dataset can be used for examining indications of learning of programming concepts and abstractions, as in \cite{Aivaloglou_2016}.  
 Existing works in this research direction include the work by Maloney \emph{et al.}~\cite{Maloney_2008}, who analyzed 536 Scratch projects created in an after-school clubhouse for blocks that relate to programming concepts and found that within the least utilized ones are Boolean operators and variables.
@@
 Another promising research direction relates to the \textit{quality of the programming artifacts developed in the Scratch environment}.
-The dataset can be used for quality assessment, for the identification of programming smells~\cite{Aivaloglou_2016}, and for exploring indications of harmful programming habits~\cite{Robles_2017}.
+The dataset has already been used for quality assessment, for the identification of programming smells~\cite{Aivaloglou_2016}, and for exploring indications of harmful programming habits~\cite{Robles_2017}.
 Existing works in this direction include a study in a classroom setting~\cite{Meerbaum_habits_2011}, which highlighted two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming.

### 2017-03-16 MSR-Scratch.tex
Shortened paper and references

@@
 
-The coding elements in the Scratch environment are the \textbf{\texttt{blocks}}, which can receive input parameters.
-In the example in the figure, block `go to x:0 y:0' receives as input two numerical parameters.
+The coding elements in the Scratch environment are the \textbf{\texttt{blocks}}, which can receive input parameters, like 
+block `go to x:0 y:0' in the figure.
 Custom clock invocations are stored in this table just like regular blocks.
@@
 The majority of the projects have been viewed only once, and they have not been favorited or remixed.
-Around 10,000 of the total projects have been remixed at least once, while the dataset includes cases of very popular projects with more than 100 remixes and even one thousand cases with more than 100 views.
+Around 10,000 of the total projects have been remixed at least once, while there are cases of very popular projects with more than 100 remixes and even one thousand cases with more than 100 views.
 
-Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21, which is considered a medium level of development of computational thinking skills.
-The table highlights the existence of surprisingly complex projects, with 525 sprites, 3,000 scripts and 34,000 blocks.
+Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21, which is considered a medium level of computational thinking skills development.
+The table highlights the existence of surprisingly complex projects, with 525 sprites and over 34,000 blocks.
 
@@
 The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education is receiving significant research attention.
-A number of studies have been carried out during the last years on understanding of the programming concepts and the programming practices of novice programmers in those environments, on the programming skills they develop, and on the quality of their programs.
+A number of studies have been carried out during the last years on understanding the programming practices of learners in those environments, on the programming skills they develop, and on the quality of their programs.
 
@@
 The authors connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops.
-Related to smell detection are the Scratch automated quality analysis tools Hairball~\cite{boe_hairball:_2013}, a lint-like static analysis tool for Scratch that can detect initialization problems and unmatched broadcast and receive blocks, and Dr.
-Scratch~\cite{moreno-leon_dr._2015}, which extends Hairball to detect two bad programming habits in Scratch: not changing the default object names and duplicating scripts.
+Related to smell detection are the Scratch automated quality analysis tools Hairball~\cite{boe_hairball:_2013}, a static analysis tool that can detect initialization problems and unmatched broadcast and receive blocks, and Dr.
+Scratch~\cite{moreno-leon_dr._2015}, which extends Hairball to detect two bad programming habits: not changing the default object names and duplicating scripts.
 
@@
 
-Enabling research towards this direction requires extending the dataset with the complete set of projects for the included users, along with their creation dates.
-This would enable examining how the programming skills and mastery of the users evolved through time, possibly through remixing the projects of other users.
-An even richer extension of the dataset would be with, possibly daily, snapshots of the included projects that are still in development.
-This would enable reconstructing some project versioning information, not available from the Scratch web interface, and thus examining how the projects are developed and evolve over time.
+Research towards this direction requires extending the dataset with the complete set of projects for the included users, along with their creation dates.
+This would enable examining how their programming skills and mastery evolved through time, possibly through remixing the projects of others.
+An even richer extension of the dataset would be with periodical snapshots of the included projects that are still in development.
+This would enable reconstructing project versioning information, not available from the Scratch interface, and thus examining how the projects are developed and evolve over time.
 

### 2017-03-16 MSR-Scratch.tex
Minor changes. For me, it is ready to be submitted.

@@
 	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
-			Email: \{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
+			\{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
 	\IEEEauthorblockA{\IEEEauthorrefmark{2}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
-		Email: jesus.moreno@programamos.es}
+		jesus.moreno@programamos.es}
 	\IEEEauthorblockA{\IEEEauthorrefmark{3}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
-		Email: grex@gsyc.urjc.es}
+		grex@gsyc.urjc.es}
 }
@@
 Some utilize program data collected during specific programming courses (e.g.,~\cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab~\cite{2017_scratch_dataset}, which contains data for Scratch projects created until 2012 (e.g.,~\cite{fields_2014, yang_2015, Dasgupta_2016}).
-In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices ~\cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr.
+In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices (e.g.,~\cite{Meerbaum_habits_2011, Aivaloglou_2016}), and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr.
 Scratch~\cite{moreno_automatic_2014}).
@@
 The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the current Scratch version (Scratch 2) and of the web programming interface in 2013.
-It is since then that the popularity of Scratch started to increase.\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}
-Scratch 2 introduced several features relating to important programming concepts,\footnote{The complete list of features introduced in Scratch 2 can be found at \url{https://wiki.scratch.mit.edu/wiki/Scratch_2.0}} like custom blocks (the equivalent to procedure definitions), and therefore that dataset does not include projects utilizing those features.
+It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
+Scratch 2 introduced several features relating to important programming concepts\footnote{The complete list of features introduced in Scratch 2 can be found at \url{https://wiki.scratch.mit.edu/wiki/Scratch_2.0}}, like custom blocks (the equivalent to procedure definitions), and therefore that dataset does not include projects utilizing those features.
 
@@
 
-We ran the scraper on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects.
+We ran the scraper on March 2\textsuperscript{nd} 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects.
 Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
@@
 		&  & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockTypes:b-type}})\\
-		&  & \texttt{parameter-1} (value of 1st input parameter)\\
+		&  & \texttt{parameter-1} (value of 1\textsuperscript{st} input parameter)\\
 		&  & ... \\
-		&  & \texttt{parameter-24} (value 24th input parameter)\\
+		&  & \texttt{parameter-24} (value 24\textsuperscript{th} input parameter)\\
 		\hline
@@
 The code is organized into \textbf{\texttt{scripts}}, i.e., groups of Scratch code blocks, each script belonging to a sprite named \texttt{sprite-name}.
-In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
+In the example in Figure~\ref{fig-scratchExample} there are three scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
 Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedures}} table.
@@
 
-Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21, which is considered a medium level of computational thinking skills development.
+Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 out of 21, which is considered a medium level of computational thinking skills development.
 The table highlights the existence of surprisingly complex projects, with 525 sprites and over 34,000 blocks.
@@
 Apart from quantitative studies, the dataset can support the design of experiments and field studies, whose material often includes Scratch programs with specific characteristics.
-For example, in~\cite{hermans_2016} examples of Scratch programs exhibiting specific smells were used in a controlled experiment to determine how the smells affect the understanding and the modification of the programs.
+For example, examples of Scratch programs exhibiting specific smells have been used in a controlled experiment to determine how the smells affect the understanding and the modification of the programs~\cite{hermans_2016}.
 This queryable dataset can support finding example programs with characteristics according to the experiment design and the field of the study.
@@
 However, other directions in the computing education research field require richer author data, and especially data on their gender and age.
-There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman~\cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelism, conditionals and, especially, variables were under-represented by all grades apart from 5 an 6.
+There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman~\cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelism, conditionals and, especially, variables were under-represented by all grades apart from 5\textsuperscript{th} and 6\textsuperscript{th}.
 The effect of gender and account age were also examined in~\cite{fields_2014} in relation to the use of programming concepts.
 %Studies on the effect of parameters like the age are not only interesting but also of direct value to educators, since they give insights on when it is better to introduce computational thinking concepts in the curriculum of schools.
-However, this dataset does not include information on the gender and the age of authors and cannot be extended to do so using the current Scratch web interface because this information is not available in the user profile page\footnotemark[\ref{fn-authorpage}], even though it is provided by users upon registration.
+However, this dataset does not include information on the gender and the age of authors, and cannot be extended to do so using the current Scratch web interface because this information is not available in the user profile page\footnotemark[\ref{fn-authorpage}], even though it is provided by users upon registration.
 
